### 2015-05-11 swmpr_manu.Rnw
minor edits before submission

@@

An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acrlong{SWMP} (\acrshort{SWMP}) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \gls{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \gls{SWMP} data.  \gls{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \gls{SWMP} \cite{SWMP14}.

@@

The \gls{CDMO} web portal was established to support priority areas of \gls{SWMP} that focus on the continuation and advancement of data management.  As such, \gls{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \gls{QAQC} procedures.    Prior to any data request to the \gls{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs. All stations in the \gls{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.  All reserve names, stations, and date ranges for each parameter type can be viewed on the \gls{CDMO} website. Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.  As noted below, the computer's IP address must be registered with \gls{CDMO} before using the data retrieval functions in SWMPr.  Web services are provided by \gls{CDMO} for direct access to \gls{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.  The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \gls{CDMO} web services.  For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \gls{SWMP} sites.  The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.

@@

The SWMPr package was developed by considering a standard workflow that categorizes the functions as one of three steps based on their intended use: retrieving, organizing, and analyzing.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes the package structure, beginning with the retrieval functions, a description of the \texttt{swmpr} object returned after retrieval, and, finally, the organizing and analyzing functions.

@@

Two approaches can be used to import \gls{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using \gls{CDMO} web services. To do so, the IP address for the computer making the request must be registered by following instructions on the \gls{CDMO} website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \gls{CDMO}.

@@

Data retrieval functions to import data directly into R from the \gls{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested. Due to rate limitations on the \gls{CDMO} server, the retrieval functions return a limited number of records with each request.  However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain the desired time series.  Download time can be excessive for longer time series.

@@

Occasionally, non-unique observations are present in the raw data.  These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data.  Erroneous duplicates with non-unique time stamps may also be present.  The \texttt{import\_local} function handles duplicate entries differently depending on the data type.  For water quality and nutrient data, duplicate time stamps are simply removed.  Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.  Nutrient data with replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.  Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.  Only duplicate values at 15 minutes are averaged for weather data during import.

@@

# this is an example path with the decompressed csv files
path <- 'C:/my_path/'

@@

# view all attributes of dat
atts <- paste0('\\texttt{', names(attributes(dat)), '}')
atts <- gsub('_', '\\_', atts, fixed = T)
atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
atts_desc <- c(
  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class',
  'Row names of the data set, inherited from the \\texttt{data.frame} object class',
  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}',
  'Station identifier used by \\gls{NERRS} as a string with 7 or 8 characters',
  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}",
  'Indicates if \\gls{QAQC} columns are present in the raw data',
  'Start and end dates for the raw data',
'Timezone of the station using the city/country format\\textsuperscript{a}',
  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated'
)

@@

The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for generic functions are developed for the object class.  A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods also apply to \texttt{swmpr} objects.  Available methods for the \texttt{swmpr} class are described below and can also be viewed:

@@

A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).  This dataset has an identical format as the data returned from the zip downloads feature of the \gls{CDMO}.  These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function.  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.  The \gls{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

descrips <- c(
  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.',
  'Remove \\gls{QAQC} columns and remove data based on \\gls{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\gls{QAQC} columns are present. ', 
  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\gls{QAQC} flags used by \\gls{CDMO}.  The output can be used to inform further processing.', 
  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \\texttt{comb} and can also be used with individual stations.',
  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
  )

@@

Raw nutrient data obtained from the \gls{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.

<<eval = F, cache = F>>=
# get nutrient data
data(apacpnut)
dat <- apacpnut
dat <- qaqc(dat)

@@

A subset method added to the existing generic \texttt{subset} function in R is available for \texttt{swmpr} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator as a character string passed to the argument, such as \texttt{`>'} or \texttt{`<='}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \texttt{`2012-11-31 12:00'}.  Finally, the function can be used to remove rows and columns that do not contain data.

@@

The \texttt{setstep} function formats a \texttt{swmpr} object to a continuous time series at a given time step.  This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.  The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.  Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.  Likewise, the default value for \texttt{differ} is one half the time step.  Time steps that do not match any existing data within the limits of the \texttt{differ} argument are not discarded, although a corresponding data value will not be assigned.

@@

The \texttt{comb} function is used to combine multiple \texttt{swmpr} objects into a single object with a continuous time series at a given step.  The \texttt{timestep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.  All arguments must be called explicitly since an arbitrary number of \texttt{swmpr} objects can be used as input.  The function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.  The time series for merging depends on the value passed to the \texttt{method} argument.  Passing \texttt{`union'} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.  Passing \texttt{`intersect'} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.  Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.  Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).

@@

The \texttt{aggreswmp} function aggregates parameter data for a \texttt{swmpr} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a predefined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.  Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).

@@

descrips <- c(
  'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of raw metabolism data.',
  'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.',
  'Decompose a \\texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \\texttt{decompose} \\cite{Kendall83}.  Decomposition of monthly or daily trends is possible.',
  'Decompose a \\texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \\texttt{decompTs} in the wq package \\cite{Jassby14}.  Only monthly decomposition is possible.',
  'Plot a histogram for a \\texttt{swmpr} object.',
  'Add lines to an existing plot created with \\texttt{plot}.',
  'Create a map of all stations in a reserve using the ggmap package.',
  'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.', 
  'Plot a univariate  time series for a \\texttt{swmpr} object.  The parameter name must be specified.',
  'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a \\texttt{swmpr} object.',  
  'Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.',
  'Smooth \\texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \\texttt{filter}.'
  )

@@

# smooth
dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')

@@

<<decomp_ex1, eval = T, fig.height = 6, cache = T, fig.cap = "An additive decomposition of dissolved oxygen into a trend, seasonal, and random component using the \\texttt{decomp} function.">>=
# get data
data(apadbwq)
swmp1 <- apadbwq

# subset for daily decomposition
dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))

# decomposition and plot
test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
plot(test)
@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.  Additionally, the time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.  For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  \Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  First, the \gls{DO} time series is assumed to be a sample of the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).

The following is an example that shows use of the function from a combined water quality and weather data set.  Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).

>>>>

% bold/italic text
\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}

@@

An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}\glsunset{NERRS}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acrlong{SWMP} (\acrshort{SWMP}) was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \gls{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \gls{SWMP} data.  \gls{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \gls{SWMP} \cite{SWMP14}.

@@

The \gls{CDMO} web portal was established to support priority areas of \gls{SWMP} that focus on the continuation and advancement of data management.  As such, \gls{CDMO} provides access to over 58 million water quality, weather, and nutrient records that have been authenticated through systematic \gls{QAQC} procedures.    Prior to any data request to the \gls{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs. All stations in the \gls{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.  All reserve names, stations, and date ranges for each parameter type can be viewed on the \gls{CDMO} website. Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.  As noted below, the computer's IP address must be registered with \gls{CDMO} before using the data retrieval functions in SWMPr.  Web services are provided by \gls{CDMO} for direct access to \gls{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.  The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \gls{CDMO} web services.  For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \gls{SWMP} sites.  The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.

@@

The SWMPr package was developed by considering a common data workflow that categorizes the functions as one of three steps based on their intended use: \Bigtxt{retrieving}, \Bigtxt{organizing}, and \Bigtxt{analyzing}.  Functions for retrieving are used to import the data into R as a \texttt{swmpr} object class.  Functions for organizing and analyzing the data provide methods for working with a \texttt{swmpr} object.  An additional group of `miscellaneous' functions are included as helpers for the main functions.  The following describes the package structure, beginning with the retrieval functions, a description of the \texttt{swmpr} object returned after retrieval, and, finally, the organizing and analyzing functions.

@@

Two basic approaches using SWMPr are available to import \gls{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using \gls{CDMO} web services. To do so, the IP address for the computer making the request must be registered by following instructions on the \gls{CDMO} website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \gls{CDMO}.

@@

Data retrieval functions to import data directly into R from the \gls{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested. Due to rate limitations on the \gls{CDMO} server, the retrieval functions return a limited number of records with each request.  However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.

@@

Occasionally, non-unique observations are present in the raw data.  These duplicates may be actual replicates with unique time stamps, such as replicate samples for monthly nutrient data, or erroneous duplicates with non-unique time stamps.  The \texttt{import\_local} function handles duplicate entries differently depending on the data type.  For water quality and nutrient data, duplicate time stamps are simply removed.  Nutrient data often contain replicate samples with similar but not identical time stamps within the span of a few minutes.  Nutrient data with replicates with unique time stamps are not removed but can be further processed using \texttt{rem\_reps}.  Weather data prior to 2007 may also contain duplicate time stamps at frequencies for hourly (denoted as `60') and daily (`144') averages, in addition to 15 minute frequencies.  Only duplicate values at 15 minutes are averaged for weather data during import.

@@

# this is an example path with the decompressed csv files
path <- 'C:/my_path'

# this is an example path for zipped csv files
path <- 'C:/my_path.zip'

@@

# view all attributes of dat
atts <- paste0('\\texttt{', names(attributes(dat)), '}')
atts <- gsub('_', '\\_', atts, fixed = T)
atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
atts_desc <- c(
  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class.',
  'Row names of the data set, inherited from the \\texttt{data.frame} object class.',
  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}.',
  'Station identifier used by \\gls{NERRS} as a string with 7 or 8 characters.',
  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}, \\texttt{`turb'}, etc.",
  'Indicates if \\gls{QAQC} columns are present in the raw data.',
  'Start and end dates for the data.',
'Timezone of the station using the city/country format\\textsuperscript{a}.',
  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.'
)

@@

The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for generic functions are developed for the object class.  A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods available in R also apply to \texttt{swmpr} objects.  Available methods for the \texttt{swmpr} class are described below and can also be viewed:

@@

A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).  This dataset has an identical format as the data returned from the zip downloads feature of the \gls{CDMO}.  These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function, as shown above.  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified \gls{QAQC} flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding \gls{QAQC} column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the \gls{QAQC} column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.  The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the \gls{QAQC} columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

descrips <- c(
  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.',
  'Remove \\gls{QAQC} columns and remove data based on \\gls{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\gls{QAQC} columns are present. ', 
  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to each \\gls{QAQC} flag used by \\gls{CDMO}.  The output can be used to inform further processing.', 
  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \\texttt{comb}.',
  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
  )

@@

Raw nutrient data obtained from the \gls{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA} values, should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.

<<eval = F, cache = F>>=
# get nutrient data, filter qaqc flags
data(apacpnut)
dat <- apacpnut
dat <- qaqc(dat)

@@

A subset method added to the existing generic \texttt{subset} function in R is available for \texttt{swmpr} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator as a character string passed to the \texttt{operator} argument, such as \texttt{`>'} or \texttt{`<='}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \texttt{`2012-11-31 12:00'}.  Finally, the function can be used to remove rows and columns that do not contain data.

@@

The \texttt{setstep} function formats a \texttt{swmpr} object to a continuous time series at a given time step.  This function is usually not necessary because most stations collect data at 15 minute intervals, but it can be used to combine data or convert an existing time series to a different interval.  The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.  Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.  Likewise, the default value for \texttt{differ} is one half the time step.  Time steps that do not match any existing data within the limits of the \texttt{differ} argument are not discarded, although a corresponding data value will not be assigned.

@@

The \texttt{comb} function is used to combine multiple \texttt{swmpr} objects into a single object with a continuous time series at a given step.  The \texttt{setstep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.  All arguments must be called explicitly since an arbitrary number of \texttt{swmpr} objects can be used as input to \texttt{comb}.  The function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.  The time series for merging depends on the value passed to the \texttt{method} argument.  Passing \texttt{`union'} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.  Passing \texttt{`intersect'} to \texttt{method} will create a continuous time series from the set of dates that are shared between all input objects.  Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.  Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).

@@

The \texttt{aggreswmp} function aggregates parameter data for a \texttt{swmpr} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.  Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).

@@

descrips <- c(
  'Aggregate \\texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.',
  'Aggregate metabolism data from a \\texttt{swmpr} object.  This is primarly used within \\texttt{plot\\_metab} but may be useful for simple summaries of daily metabolism data.',
  'Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.',
  'Decompose a \\texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \\texttt{decompose} \\cite{Kendall83}.  Decomposition of monthly or daily trends is possible.',
  'Decompose a \\texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \\texttt{decompTs} in the wq package \\cite{Jassby14}.  Only monthly decomposition is possible.',
  'Plot a histogram for a \\texttt{swmpr} object.',
  'Add lines to an existing plot created with \\texttt{plot}.',
  'Create a map of all stations in a reserve using the ggmap package \\cite{Kahle13}.',
  'Linearly interpolate missing data (\\texttt{NA} values) in a \\texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.', 
  'Plot a univariate  time series for a \\texttt{swmpr} object.  The parameter name must be specified.',
  'Plot ecosystem metabolism estimates after running \\texttt{ecometab} on a combined \\texttt{swmpr} object.',  
  'Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.',
  'Smooth \\texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \\texttt{filter}.'
  )

@@

# smooth using a window of 50 observations
dat_smooth <- smoother(dat, window = 50, params = 'do_mgl')

@@

<<decomp_ex1, eval = T, fig.height = 6, cache = T, fig.cap = "An additive decomposition of dissolved oxygen into a trend, seasonal (daily), and random component using the \\texttt{decomp} function.">>=
# get data
data(apadbwq)
dat <- apadbwq

# subset for daily decomposition
dat <- subset(dat, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))

# daily decomposition of DO and plot
dc_dat <- decomp(dat, param = 'do_mgl', frequency = 'daily')
plot(dc_dat)
@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.  The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.  For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  \Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  First, the \gls{DO} time series is assumed to be a sample of the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a substantial amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).

The following is an example that shows use of the \texttt{ecometab} function from a combined water quality and weather data set.  Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).
